Architectures for workflow integration in science gateways

### 2016-04-18 paper.tex
Major update in section 4 (metrics have changed).

\section{Architectures}

The architectures are diagramed in Figure~\ref{fig:architectures}
using the graphical notations shown in
Figure~\ref{fig:notations}. Architectures are described from their
main software \emph{components} and \emph{interactions}. Software
components include science gateway and infrastructure in all
architectures, and workflow service, workflow pool and agent are
involved in some architectures. The workflow engine itself is
represented by a specific symbol. Software interactions among these
components are described in the next Section. \emph{Abstract}
interactions are specific types of interactions that may be
implemented by various different software interactions. An
architecture that has an abstract interaction is an abstract
architecture. Red color is used to represent the components and
interactions that need to be specifically developed to integrate
workflow engines. Tasks refer to computing tasks that are created by
the workflow engine and executed on the infrastructure. Data
represents any type of file, or database that is involved in the
workflow execution and stored on the infrastructure while the workflow
executes. It does not cover workflow parameters such as strings,
numbers, etc. The reminder of this section describes the different
types of interactions involved in the architectures, and details each
architecture.

@@

\todo{In the following sections, for each architecture, add a
  description of a real system that implements it: integrated: ?,
  service: VIP, sub-task: CBRAIN-PSOM, pool: SHIWA pool, nested
  workflow: SHIWA CGI, import: SHIWA FGI}

@@

See Figure~\ref{archi:tight}. The workflow engine is tightly
integrated with the science gateway, which means that it is deployed
on the same machine and potentially shares code, libraries and other
software components with the science gateway. For instance, the
workflow engine might be a portlet in a Liferay portal or a controller
in a Ruby on Rails application. The workflow engine and the science
gateway usually share a database where application, users and other
resources are stored. In this model, task and data control are both
initiated from the science gateway. Interactions \texttt{b} and
\texttt{c$_2$} are initiated from the workflow engine while
\texttt{c$_1$} comes from other parts of the science gateway, for
instance a data management interface. As in any other model, the
installation of new workflows in the science gateway (\texttt{a$_2$})
and infrastructure (\texttt{a$_1$}) is done by an administrator, for
security reasons. This is the model adopted in the Catania Science
Gateway Framework~\cite{Ardizzone2012} (see specific documentation on
workflows\footnote{\url{http://bit.ly/1oQrzvQ}}),
CIPRES~\cite{miller2010creating} and LONI Pipeline
Environment~\cite{dinov2009efficient}.

@@

\todo{workflow import is an offline process.}

@@

See Table~\ref{table:evaluation}. We focus on the integration of
workflow engines in science gateways, i.e. we assume that the science
gateway is already interfaced with the infrastructure.

@@

We use four main criteria to evaluate the architectures: integration
effort, robustness, modularity and scalability. Criteria break down to
specific metrics for which low values indicate good performance.

@@

\todo{Refer to the metric codes.}

\emph{Integration} measures the development effort required to build
the architecture, assuming that the science gateway and infrastructure
are already integrated. It is measured by counting on the
architectural diagram the number of interactions and components to
develop or modify. It breaks down to 4 metrics that quantify the
developments related to:
\begin{itemize}[itemsep=0cm]
\item Science gateway (\texttt{I$_1$}),
\item Workflow engine (\texttt{I$_2$}),
\item Infrastructure (\texttt{I$_3$}),
\item Other components (\texttt{I$_4$}).
\end{itemize}
Interactions are counted only once, in the component from which they
originate.

\emph{Robustness} relates to the complexity of the architecture and
its ability to provide relevant debugging information about the
workflow executions. It breaks down to 3 metrics:
\begin{itemize}[itemsep=0cm]
\item Components: number of unique components involved in workflow executions.
\item Interactions: number of unique interactions involved in workflow executions. 
\item Debugging: availability of debugging information about workflow
  activities (1: difficult to obtain; 0: easy to access). Fine-grained
  information about workflow activity is required to properly
  troubleshoot workflow executions.
\end{itemize}

\emph{Modularity} measures the ability to replace elements in the
architecture, or to integrate new elements. Elements include
workflow engines, workflows and infrastructure. It breaks down in 4
metrics:
\begin{itemize}[itemsep=0cm]
\item New engine type: number of interactions or components to modify
  to integrate a new workflow engine type in the
  architecture. Modification of an interaction is deemed necessary
  when the parameters involved in this interaction are
  modified. Modification of a component is required when its code
  needs to be modified or recompiled (science gateway or workflow
  service), or when a new piece of software has to be installed
  (infrastructure only). Adding a new type of workflow engine allows
  to execute the workflows that have been described in the language(s)
  used by the new engine.
\item New engine version: number of interactions or components to
  modify to integrate a new version of a workflow engine in the
  architecture, assuming that another version of the same engine type
  is already available. We assume that different versions of a
  workflow engine share the same interface, i.e. they can be invoked
  using the same software. When this is not the case, the different
  versions have to be considered as different engine types. 
\item New infrastructure: number of interactions or components to
  modify to integrate a new type of infrastructure in the
  architecture.
\item New workflow: number of interactions required to integrate a new
  workflow in the architecture.
\item Meta-workflow: ability to describe meta-workflows from existing
  workflows (1: not available; 0: available).
\end{itemize}

\emph{Scalability} measures the ability of the architecture to support
high workloads. It breaks down into 4 metrics:
\begin{itemize}[itemsep=0cm]
\item New engine instance (0: automated; 1: manual intervention needed; 2: not possible).
\item Elastic engines (0: easy to implement; 1: difficult to
  implement; 2: not available).
\item Distributed engines (0: possible; 1: not possible).
\item Task scheduling (0: difficult; 1: more difficult).
\end{itemize}

@@

@@

\paragraph{Integration} Tightly integrating a workflow engine in a
science gateway requires to modify the science gateway
(\texttt{I$_1$=1}) and to develop interactions \texttt{b}, and
\texttt{c$_2$} in the workflow engine (\texttt{I$_2$=2}). No other
modification is required (\texttt{I$_3$}=\texttt{I$_4$}=0).

\paragraph{Robustness} The science gateway and the infrastructure are
involved in a workflow execution (\texttt{R$_1$}=2), with interactions
\texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
(\texttt{R$_2$}=3). Debugging is not an issue since the science
gateway can retrieve any information from the workflow engine
directly (\texttt{R$_3$}=0).

\paragraph{Modularity} Integrating a new type of workflow engine
requires to modify the science gateway as well as interactions
\texttt{b} and \texttt{c$_2$} (\texttt{M$_1$}=3). Updating a workflow
engine version requires modifications in the science gateway
(\texttt{M$_2$}=1). Adding a new infrastructure generates updates in
interactions \texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
(\texttt{M$_3$}=3). Inserting a new workflow is done through
interaction \texttt{a} (\texttt{M$_4$}=1). Meta-workflows are not
supported by default (\texttt{M$_4$}=1).

\paragraph{Scalability} Adding a new engine instance requires a new
instance of the science gateway, which is in general not possible
since load-balancing mechanisms available in web servers are usually
meant to deal with short web requests rather than long workflow
executions (\texttt{S$_1$}=2). Coherently, elastic engines are usually
not possible either (\texttt{S$_2$}=2).  Distributed engines are not
available by default (\texttt{S$_3$}=1). The scheduling of tasks on
the infrastructure is as complex as in any other architecture since
the workflow engine might implement any kind of scheduling policy
(\texttt{S$_4$}=0).

@@

\paragraph{Integration} Integrating a workflow engine as a service
requires to develop 3 interactions and 1 component. First, a workflow
service has to be developed and interfaced with the infrastructure
through interactions \texttt{b} and \texttt{c$_2$}
(\texttt{I$_2$}=3). In addtion, the science gateway has to be extended
to call the engine interface \texttt{d} (\texttt{I$_1$}=1).

\paragraph{Robustness} Workflow execution requires 3 components: the
science gateway, the workflow service, and the infrastructure
(\texttt{R$_1$}=3). It involves the 4 interactions \texttt{b},
\texttt{$c_1$}, \texttt{$c_2$} and \texttt{d}
(\texttt{R$_2$}=4). Fine-grained debugging information is usually easy
to obtain since the workflow service provides direct access to the
engine (\texttt{R$_3$}=0).

\paragraph{Modularity} Adding a new type of workflow engine requires
to implement the corresponding workflow service, to modify interaction
\texttt{d}, and to implement interactions \texttt{b} and
\texttt{$c_2$} (\texttt{M$_1$}=4). New engine versions can be added by
updating the workflow service without modifying any interaction
(\texttt{M$_2$}=0). Adding a new type of infrastructure requires
updates in interactions \texttt{b}, \texttt{$c_1$} and \texttt{$c_2$}
(\texttt{M$_3$}=3). New workflows are added in the science gateway or
in the workflow engine through interaction \texttt{a} only
(\texttt{M$_4$}=1). Meta-workflow are not supported by default
(\texttt{M$_5$}=1).

\paragraph{Scalability} The service architecture supports multiple
engine instances through multiple workflow services. In VIP for
instance, this feature has been available from release 1.17 (April
2016). A basic load-balancing mechanism is available that sends new
workflow executions to the engine instance that has the least active
executions. To avoid ``black-hole'' syndromes created by failing
engine instances, engine instances are automatically disabled when
workflows cannot be submitted to them. Adding a new engine instance,
however, requires manual intervention to declare the new instance in
the science gateway (\texttt{S$_1$}=1). Consequently, elastic engines
are difficult to implement because they require a mechanism to update
the science gateway configuration when a new engine instance is
available (\texttt{S$_2$}=1). Distributing the execution of a single
workflow in multiple engines is usually not possible unless the
workflow engine has specific abilities (\texttt{S$_3$}=1). The
scheduling of tasks on the infrastructure is as complex as in any
other architecture since the workflow engine might implement any kind
of scheduling policy (\texttt{S$_4$}=0).

@@

\paragraph{Integration} Integrating a workflow engine as a sub-task
only requires to implement interaction \texttt{e} (\texttt{I$_2$}=1)
and to install the workflow engine on the infrastructure
(\texttt{I$_3$}=1). A new task type is created in the science gateway
only when a new workflow has to be integrated (see M$_4$), but this is
not required to integrate the engine itself (\texttt{I$_1$}=0).

\paragraph{Robustness} Only 2 components and 3 interactions are
involved in workflow execution (\texttt{R$_1$}=2,
\texttt{R$_2$}=3). Obtaining fine-grained information about workflow
activities is not straightforward since the science gateway has no
knowledge about the workflow topology, and the workflow engine is
integrated as a task (\texttt{R$_3$}=1).

\paragraph{Modularity} Integrating a new type of workflow engine
requires to develop interaction \texttt{e} and to install the engine
on the infrastructure (\texttt{M$_1$}=2). Updating an engine version
requires updates only on the infrastructure (\texttt{M$_2$}=1). Adding
a new infrastructure requires to update interactions \texttt{b} and
\texttt{c} in the science gateway (\texttt{M$_3$}=2). New workflows are integrated by
creating a new task in the science gateway through interaction
\texttt{a} (\texttt{M$_4$}=1), and meta-workflows are not available
(\texttt{M$_5$}=1).

\paragraph{Scalability}
New engine instances are spawned and executed on the infrastructure as
any other task upon user submission (\texttt{S$_1$}=0,
\texttt{S$_2$}=0). Such scalability properties are one of the major
interests of the sub-task architecture. Distributed engines are not
supported by default (\texttt{S$_3$}=1). Task scheduling is slightly
more complex than in the other approaches due to the special role of
the task that executes the workflow engine (\texttt{S$_4$}=1). Indeed,
the reliability of this task is critical since all the sub-tasks in
the workflow depend on it and, depending on the recovery capabilities
of the workflow engine, may need to be resubmitted if the workflow
task fails. The workflow task is also longer than all its sub-tasks,
which increases its chances of failure. In addition, task parameters,
for instance estimated walltime, are more difficult to estimate for
the workflow task than for sub-tasks which may generate issues such as
selection of wrong batch queues on clusters. Finally, the
interdependencies between the workflow task and its sub-tasks may
create deadlocks when there is contention. Say for instance that only
1 computing resource is available for the science gateway and that the
workflow task is running on it and submits sub-tasks, then the
sub-tasks could only execute when the resource is available, which
will never happen because the workflow task will not complete until
the sub-tasks complete. This configuration can be generalized to an
infrastructure with $n$ resources where $n$ workflows are
submitted. In practice, however, the number of submitted workflows
usually remains lower than the number of computing resources available
on this infrastructure, which makes such deadlocks unlikely to happen.

@@

\paragraph{Integration} Integrating workflow engines through the pool
model requires interaction \texttt{d} in the science gateway
(\texttt{I$_1$=1}), interactions \texttt{b} and \texttt{c$_2$} in the
workflow engine (\texttt{I$_2$=2}), the development of the workflow
pool, agent and interaction \texttt{f} between them
(\texttt{I$_4$=3}).

\paragraph{Robustness} The science gateway, workflow pool, agent and
infrastructure are involved in a workflow execution
(\texttt{R$_1$=4}), and these components are connected through 4
interactions (\texttt{R$_2$=4}). Accessing debugging information is
not likely to be an issue since the workflow pool could implement
specific functions for that (\texttt{R$_3$=0}).

\paragraph{Modularity} Adding a new engine type requires to wrap the
engine in the agent and to update interactions \texttt{b} and
\texttt{c$_2$} (\texttt{M$_1$=3}). Updating the version of an engine
is transparent (\texttt{M$_2$=0}), and integrating a new
infrastructure requires updates in interactions \texttt{b},
\texttt{c$_1$} and \texttt{c$_2$} (\texttt{M$_3$=3}). Integrating a
new workflow is done through interaction \texttt{a} only
(\texttt{M$_4$=1}) and meta-workflows are not available by default
(\texttt{M$_5$=1}).

\paragraph{Scalability} New engine instances only require new agents,
which is easily automated (\texttt{S$_1$=0}) and by design very
suitable for elastic computing (\texttt{S$_2$=0}). For instance,
auto-scaling rules can be implemented to start new agents when the
workload in the science gateway exceeds a certain threshold \todo{ref
  needed}. Distributed engines are not available by default
(\texttt{S$_3$=1}) and task scheduling is as complex as in any other
architecture (\texttt{S$_4$=0}).

@@

\paragraph{Integration} Setting up a nested workflow architecture with
service invocation requires interaction \texttt{d} in the science
gateway (\texttt{I$_1$=1}), and the development of 2 workflow services
with 2 instances of interactions \texttt{b} and \texttt{c$_2$}
(\texttt{I$_2$=5}).

\paragraph{Robustness} Workflow execution involves the science
gateway, 2 workflow services and the infrastructure
(\texttt{R$_1$=4}), which includes 7 interactions
(\texttt{R$_2$=7}). Debugging is difficult because the science gateway
cannot directly access fine-grained information in the sub-workflow
(\texttt{R$_3$=1}).

\paragraph{Modularity} Adding a new type of \emph{parent} engine
requires to implement the corresponding service, to implement
interactions \texttt{b} and \texttt{c$_2$} in the parent engine, and
to implement interaction \texttt{d} in the science gateway and in the
parent service (\texttt{M$_1$=5}). Adding a new type of \emph{child}
engine only requires to implement the corresponding service, to
develop interactions \texttt{b} and \texttt{c$_2$} in the child
engine, and to implement interaction \texttt{d} in the parent service
(\texttt{M$_1$=4}). We use \texttt{M$_1$=4.5} in
Table~\ref{table:evaluation} to reflect both conditions. Adding a new
version in the parent or child engine only requires modifying this
engine (\texttt{M$_2$=0}). Adding a new infrastructure requires to
re-implement interactions \texttt{b} and \texttt{c$_2$} twice, and
interaction \texttt{c$_1$} once (\texttt{M$_3$=5}). Adding a new
workflow is done through interaction \texttt{a}
(\texttt{M$_4$=1}). Meta-workflows are possible, which is one of the
main interest of this architecture (\texttt{M$_5$=0}).

>>>>

\section{Architectures}
\label{sec:architectures}

The architectures are diagramed in Figure~\ref{fig:architectures}
using the graphical notations shown in
Figure~\ref{fig:notations}. Architectures are described from their
main software \emph{components} and \emph{interactions}. Software
components include science gateway and infrastructure in all
architectures, and workflow service, workflow pool and agent are
involved in some architectures. The workflow engine itself is
represented by a specific symbol. Software interactions among these
components are described in the next Section. \emph{Abstract}
interactions are specific types of interactions that may be
implemented by various different software interactions. An
architecture that has an abstract interaction is an abstract
architecture. Red color is used to represent the components and
interactions that are specific to workflow execution, i.e. that are
not involved in any other process. Tasks refer to computing tasks that
are created by the workflow engine and executed on the
infrastructure. Data represents any type of file, or database that is
involved in the workflow execution and stored on the infrastructure
while the workflow executes. It does not cover workflow parameters
such as strings, numbers, etc. The remainder of this section describes
the different types of interactions involved in the architectures, and
details each architecture.

@@

@@

See Figure~\ref{archi:tight}. The workflow engine is tightly
integrated with the science gateway, which means that it is deployed
on the same machine and potentially shares code, libraries and other
software components with the science gateway. For instance, the
workflow engine might be a portlet in a Liferay portal or a controller
in a Ruby on Rails application. The workflow engine and the science
gateway usually share a database where application, users and other
resources are stored. In this model, task and data control are both
initiated from the science gateway. Interactions \texttt{b} and
\texttt{c$_2$} are initiated from the workflow engine while
\texttt{c$_1$} comes from other parts of the science gateway, for
instance a data management interface. As in any other model, the
installation of new workflows in the science gateway (\texttt{a$_2$})
and infrastructure (\texttt{a$_1$}) is done by an administrator, for
security reasons. This is the model adopted in the Catania Science
Gateway Framework~\cite{Ardizzone2012} (see specific documentation on
workflows\footnote{\url{http://bit.ly/1oQrzvQ}}),
CIPRES~\cite{miller2010creating} and LONI Pipeline
Environment~\cite{dinov2009efficient}.
\todo{Add description of a real system.}

@@

\todo{workflow import is an offline process.}
\todo{Add description of a real system}

@@

The architectures described in Section~\ref{sec:architectures} are
evaluated on Table~\ref{table:evaluation}. We use 5 main criteria to
evaluate the architectures: integration effort, robustness of workflow
execution, extensibility, scalability and other features. Criteria
break down to specific metrics where \emph{low value indicates good
  performance}. For each criterion, a total score is computed by
summing up the individual metrics. We ensure that the different
metrics in a criterion measure similar entities so that summation
makes sense. Metric values obtained by each architecture are
normalized between 0 (best value) and 1 (worst value) to determine the
color of the corresponding table cell (brighter color indicate better
performance\footnote{Normalized metric value $m'$ is
  $\frac{m-m_{\mathrm{min}}}{m_{\mathrm{min}}-m_{\mathrm{max}}}$ where
  $m$ is the initial metric value,
  $m_{\mathrm{min}}$/$m_{\mathrm{max}}$ are the minimal/maximal values
  among all architectures. Cell color is \todo{\#99XX99, where
    X=F-4m'})}. The criteria and metrics are explained hereafter.

@@

@@

\emph{Integration} measures the development effort required to build
the architecture. It is obtained by counting the number of
interactions and components on the architectural diagram. It breaks
down to the following 2 metrics:
\begin{itemize}[itemsep=0cm]
\item Components \texttt{I$_1$}),
\item Interactions (\texttt{I$_2$}).
\end{itemize}
One may wonder whether infrastructure should be part of
\texttt{I$_1$}) since it is usually not developed by the groups who
integrate workflow engines in science gateway. However, integrating an
infrastructure in the architecture usually requires some technical
effort (e.g. account creation, software installation, etc) which is
why we keep it in the metric. The two metrics are summed to obtain
the total score for this criterion, which measures the total number of
software pieces to develop.

\emph{Robustness of workflow execution} measures the likelihood that
workflow execution fails due to errors coming from the software
architecture. It is determined as the number of software components and
interactions that are specific to workflow execution. It assume that
complex architectures tend to be prone to failure.  These components
and interactions are represented in red on the architecture diagram in
Figure~\ref{fig:architectures}. Robustness breaks down to 2 metrics:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Components (\texttt{R$_1$}): number of specific components
  involved in workflow executions.
\item Interactions (\texttt{R$_2$}): number of specific interactions
  involved in workflow executions.
\end{itemize}
The two metrics are summed to obtain the total score for this
criterion, which measures the total number of software pieces that are
specifically involved in workflow execution. Robustness of workflow
execution is critical since it directly impacts the end users of the
science gateway.

\emph{Extensibility} measures the difficulty to replace or add
elements in the architecture. It is obtained as the number of
interaction and component to modify when a new element is
added. Modification of an interaction is deemed necessary when the
parameters involved in this interaction are modified. Modification of
a component is required when its code needs to be modified or
recompiled (science gateway or workflow service), or when a new piece
of software has to be installed (infrastructure only). Extensibility
breaks down in 4 metrics depending on the type of element that has to
be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine type (\texttt{E$_1$}): number of interactions or
  components to modify to integrate a new workflow engine type in the
  architecture. Workflow engine belong to different types when they
  cannot be invoked using the same interface. Adding a new type of
  workflow engine allows to execute more workflows in the science gateway.
\item New engine version (\texttt{E$_2$}): number of interactions or
  components to modify to integrate a new version of a workflow engine
  in the architecture, assuming that another version of the same
  engine type is already available. Different versions of a workflow
  engine share the same interface, i.e. they can be invoked using the
  same software. When this is not the case, the different versions
  have to be considered as different engine types. 
\item New workflow (\texttt{E$_3$}): number of interactions required
  to integrate a new workflow in the architecture, assuming that the
  engine type and version required to execute this workflow are
  already available.
\item New infrastructure (\texttt{E$_4$}): number of interactions or components to
  modify to integrate a new type of infrastructure in the
  architecture. Adding a new infrastructure allows to provide more
  computing or storage power, to access specific types of resources
  (e.g. GPUs, clouds), or to enforce execution policies
  (e.g. constrain data to remain in a particular network domain).
\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The 4 metrics are summed to obtain a global index that measures the
difficulty to extend the architecture.

\emph{Scalability} corresponds to the ability of the architecture to
cope with high workloads. It is measured by reviewing the features
that enable particular aspects of scalability. Features are evaluated
using a 3-level metric: 0 means that the feature is very easy to
enable, 1 means that it can be implemented but with some difficulty,
and 2 means that it is not possible to implement the feature without a
nonsensical amount of effort. Four different features are identified:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Adding engine instance (\texttt{S$_1$}): this feature
  corresponds to the possibility to add more than 1 engine instance in
  the architecture. Workflow engines may require important amounts of
  resources when a lot of workflows, or large workflows, are
  executed. At some point, it may be required for the science gateway
  to distribute the load to several engines. \texttt{S$_1$}=0 when
  adding a new engine instance is a fully automated process, i.e. the
  workflow engine only has to be started. In this case, elastic
  engines are possible, i.e. some kind of auto-scaling mechanism can
  be implemented to control the number of engine instances in the
  architecture. \texttt{S$_1$}=1 when adding a new engine instance
  requires some form of manual intervention, which prevents easy
  implementations of elastic engines. \texttt{S$_1$}=2 when new engine
  instances cannot be added.
\item Distributed engines (\texttt{S$_2$}): possibility to distribute
  the execution \emph{of a single workflow} among different engine
  instances. In our scope, this feature focuses on the capabilities of
  the architecture rather than these of the workflow
  engine. \texttt{S$_2$}=0 when distributed engines are possible in
  the architecture and \texttt{S$_2$}=1 when they require specific
  developments in the workflow engine.
\item Task scheduling (\texttt{S$_3$}): task scheduling is a difficult
  issue that depends more on the implementation of specific algorithms
  in the science gateway, workflow engine and infrastructure than on
  the architecture used to integrate the workflow engine in the
  science gateway. Some architectures, however, complexify the task
  scheduling problem by introducing additional software layers or
  create tasks with specific characteristics. \texttt{S$_3$}=0 when
  the architecture does not introduce any additional complexity to the
  scheduling problem and \texttt{S$_3$}=1 otherwise.
\end{itemize}
The 3 metrics are summed to obtain a global measure of the scalability
issues available in the architecture.

\emph{Other features} include:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Meta-workflow (\texttt{O$_1$}): ability to describe meta-workflows from existing
  workflows.  \todo{More details about why this is important?} \texttt{O$_1$}=0 when the feature is not available, \texttt{O$_1$}=1 otherwise.
\item Fine-grained debugging (\texttt{O$_2$}): availability of
  debugging information about workflow activities. Fine-grained
  information about workflow activity is required to properly
  troubleshoot workflow executions.  \texttt{O$_2$}=0 when such
  information is easy to access, \texttt{O$_2$}=1
  otherwise. \todo{More details about why this is important?}
\end{itemize}
These 2 metrics are summed to obtain a total number of missing other
features in the architecture.  \todo{Maybe we could revise the
  categories to avoid such a vague ``other features''.}

@@

The different architectures described in
Figure~\ref{fig:architectures} are evaluated along these metrics in
the remainder of this Section.

@@

\paragraph{Integration} This architecture does not require any
component in addition to the science gateway and infrastructure
(\texttt{I$_1$=2}). It involves 5 interactions: \texttt{a$_1$},
\texttt{a$_2$}, \texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
(\texttt{I$_2$=5}).

\paragraph{Robustness} No component is specific to workflow execution
(\texttt{R$_1$}=0), but interactions \texttt{b} and \texttt{c$_2$}
(\texttt{R$_2$}=2) are.

\paragraph{Extensibility} Integrating a new type of workflow engine
requires to modify the science gateway as well as interactions
\texttt{b} and \texttt{c$_2$} (\texttt{E$_1$}=3). Updating a workflow
engine version requires modifications in the science gateway
(\texttt{E$_2$}=1).  Inserting a new workflow is done through
interaction \texttt{a} (\texttt{E$_3$}=1). Adding a new infrastructure generates updates in
interactions \texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
(\texttt{E$_4$}=3).

\paragraph{Scalability} Adding a new engine instance requires a new
instance of the science gateway, which is in general not possible
since load-balancing mechanisms available in web servers are usually
meant to deal with short web requests rather than long workflow
executions (\texttt{S$_1$}=2). Distributed engines are not
available by default (\texttt{S$_2$}=1). The scheduling of tasks on
the infrastructure is as complex as in any other architecture since
the workflow engine might implement any kind of scheduling policy
(\texttt{S$_3$}=0).

\paragraph{Other features} Meta-workflows are not supported by default
(\texttt{O$_1$}=1).  Debugging is not an issue since the science
gateway can retrieve any information from the workflow engine directly
(\texttt{O$_2$}=0).

@@

\paragraph{Integration} Service invocation requires a workflow service
in addition to the science gateway and infrastructure
(\texttt{I$_1$}=3). The architecture involves 6 interactions:
\texttt{a$_1$}, \texttt{a$_2$}, \texttt{b}, \texttt{c$_1$},
\texttt{c$_2$} and \texttt{d} (\texttt{I$_2$}=6).

\paragraph{Robustness} The workflow service is a component specific to
workflow execution (\texttt{R$_1$}=1). Workflow execution also
involves 3 specific interactions: \texttt{b}, 
\texttt{c$_2$} and \texttt{d} (\texttt{R$_2$}=3).

\paragraph{Extensibility} Adding a new type of workflow engine requires
to implement the corresponding workflow service, to modify interaction
\texttt{d}, and to implement interactions \texttt{b} and
\texttt{$c_2$} (\texttt{E$_1$}=4). New engine versions can be added by
updating the workflow service without modifying any interaction
(\texttt{E$_2$}=0).  New workflows are added in the science gateway or
in the workflow engine through interaction \texttt{a} only
(\texttt{E$_3$}=1). Adding a new type of infrastructure requires
updates in interactions \texttt{b}, \texttt{$c_1$} and \texttt{$c_2$}
(\texttt{E$_4$}=3).

\paragraph{Scalability} The service architecture supports multiple
engine instances through multiple workflow services. In VIP for
instance, this feature has been available from release 1.17 (April
2016). A basic load-balancing mechanism is available that sends new
workflow executions to the engine instance that has the least active
executions. To avoid ``black-hole'' syndromes created by failing
engine instances, engine instances are automatically disabled when
workflows cannot be submitted to them. Adding a new engine instance,
however, requires manual intervention to declare the new instance in
the science gateway (\texttt{S$_1$}=1). Consequently, elastic engines
are difficult to implement because they require a mechanism to update
the science gateway configuration when a new engine instance is
available. Distributing the execution of a single
workflow in multiple engines is usually not possible unless the
workflow engine has specific abilities (\texttt{S$_2$}=1). The
scheduling of tasks on the infrastructure is as complex as in any
other architecture since the workflow engine might implement any kind
of scheduling policy (\texttt{S$_3$}=0).

\paragraph{Other features} Meta-workflow are not supported by default
(\texttt{O$_1$}=1). Fine-grained debugging information is usually easy
to obtain since the workflow service provides direct access to the
engine (\texttt{O$_2$}=0).

@@

\paragraph{Integration} The sub-task architecture requires only 2
components (\texttt{I$_1$}=2).  It involves 5 interactions:
\texttt{a$_1$}, \texttt{a$_2$}, \texttt{b}, \texttt{c} and
\texttt{e} (\texttt{I$_2$}=5).

\paragraph{Robustness} No component is specific to workflow execution
(\texttt{R$_1$}=0), and only interaction \texttt{e} is
(\texttt{R$_2$}=1).

\paragraph{Extensibility} Integrating a new type of workflow engine
requires to develop interaction \texttt{e} and to install the engine
on the infrastructure (\texttt{E$_1$}=2). Updating an engine version
requires updates only on the infrastructure (\texttt{E$_2$}=1).  New
workflows are integrated by creating a new task in the science gateway
through interaction \texttt{a} (\texttt{E$_3$}=1). Adding a new
infrastructure requires to update interactions \texttt{b} and
\texttt{c} in the science gateway (\texttt{E$_4$}=2).

\paragraph{Other features} Meta-workflows are not available
(\texttt{O$_1$}=1).  Obtaining fine-grained information about workflow
activities is not straightforward since the science gateway has no
knowledge about the workflow topology, and the workflow engine is
integrated as a task (\texttt{O$_2$}=1).

%\todo{Integrating a workflow engine as a sub-task
%only requires to implement interaction \texttt{e} (\texttt{I$_2$}=1)
%and to install the workflow engine on the infrastructure
%(\texttt{I$_3$}=1). A new task type is created in the science gateway
%only when a new workflow has to be integrated (see M$_4$), but this is
%not required to integrate the engine itself (\texttt{I$_1$}=0).}

\paragraph{Scalability}
New engine instances are spawned and executed on the infrastructure as
any other task upon user submission (\texttt{S$_1$}=0). Such
scalability properties are one of the major interests of the sub-task
architecture. Distributed engines are not supported by default
(\texttt{S$_2$}=1). Task scheduling is slightly more complex than in
the other approaches due to the special role of the task that executes
the workflow engine (\texttt{S$_3$}=1). Indeed, the reliability of
this task is critical since all the sub-tasks in the workflow depend
on it and, depending on the recovery capabilities of the workflow
engine, may need to be resubmitted if the workflow task fails. The
workflow task is also longer than all its sub-tasks, which increases
its chances of failure. In addition, task parameters, for instance
estimated walltime, are more difficult to estimate for the workflow
task than for sub-tasks which may generate issues such as selection of
wrong batch queues on clusters. Finally, the interdependencies between
the workflow task and its sub-tasks may create deadlocks when there is
contention. Say for instance that only 1 computing resource is
available for the science gateway and that the workflow task is
running on it and submits sub-tasks, then the sub-tasks could only
execute when the resource is available, which will never happen
because the workflow task will not complete until the sub-tasks
complete. This configuration can be generalized to an infrastructure
with $n$ resources where $n$ workflows are submitted. In practice,
however, the number of submitted workflows usually remains lower than
the number of computing resources available on this infrastructure,
which makes such deadlocks unlikely to happen.

@@

\paragraph{Integration} The pool model requires a workflow pool and an
agent in addition to the science gateway and infrastructure
(\texttt{I$_1$=4}). It involves 7 interactions: \texttt{a$_1$},
\texttt{a$_2$}, \texttt{b}, \texttt{c$_1$}, \texttt{c$_2$}, \texttt{d}
and \texttt{f} (\texttt{I$_2$}=7).

\paragraph{Robustness} The workflow pool and agent are specific to
workflow execution (\texttt{R$_1$=2}). Interactions \texttt{b},
\texttt{c$_2$}, \texttt{d} and \texttt{f} also are (\texttt{R$_2$=4}).

\paragraph{Extensibility} Adding a new engine type requires to wrap
the engine in the agent and to update interactions \texttt{b} and
\texttt{c$_2$} (\texttt{E$_1$=3}). Updating the version of an engine
is transparent (\texttt{E$_2$=0}), and integrating a new workflow is
done through interaction \texttt{a} only
(\texttt{E$_3$=1}). Integrating a new infrastructure requires updates
in interactions \texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
(\texttt{E$_4$=3}).

\paragraph{Scalability} New engine instances only require new agents,
which is easily automated (\texttt{S$_1$=0}) and by design very
suitable for elastic computing. For instance, auto-scaling rules can
be implemented to start new agents when the workload in the science
gateway exceeds a certain threshold \todo{ref needed}. Distributed
engines are not available by default (\texttt{S$_2$=1}) and task
scheduling is as complex as in any other architecture
(\texttt{S$_3$=0}).

\paragraph{Other features} Meta-workflows are not available by default
(\texttt{O$_1$=1}).  Accessing debugging information is not likely to
be an issue since the workflow pool could implement specific functions
for that (\texttt{O$_2$=0}).

@@

\paragraph{Integration} Setting up a nested workflow architecture with
service invocation requires a science gateway, 2 workflow services and
2 infrastructures (\texttt{I$_1$=5}). The architecture involves 11
interactions (\texttt{I$_2$=11}): \texttt{a$_1$} (twice: once for each
infrastructure), \texttt{a$_2$}, \texttt{b} (twice: once for each
infrastructure), \texttt{c$_1$} (twice: once for each infrastructure),
\texttt{c$_2$} (twice: once for each workflow engine and
infrastructure) and \texttt{d} (twice: once for each workflow engine).

\paragraph{Robustness} The two workflow services are specific to
workflow execution (\texttt{R$_1$=2}). Interactions \texttt{b}
(counted twice), \texttt{c$_2$} (twice) and \texttt{d} (twice) also
are (\texttt{R$_2$=6}).

\paragraph{Extensibility} Adding a new type of \emph{parent} engine
requires to implement the corresponding service, to implement
interactions \texttt{b} and \texttt{c$_2$} in the parent engine, and
to implement interaction \texttt{d} in the science gateway and in the
parent service (\texttt{E$_1$=5}). Adding a new type of \emph{child}
engine only requires to implement the corresponding service, to
develop interactions \texttt{b} and \texttt{c$_2$} in the child
engine, and to implement interaction \texttt{d} in the parent service
(\texttt{E$_1$=4}). We use \texttt{E$_1$=4.5} in
Table~\ref{table:evaluation} to reflect both conditions. Adding a new
version in the parent or child engine only requires modifying this
engine (\texttt{E$_2$=0}). Adding a new
workflow is done through interaction \texttt{a}
(\texttt{E$_3$=1}). Adding a new infrastructure requires to
re-implement interactions \texttt{b} and \texttt{c$_2$} twice, and
interaction \texttt{c$_1$} once (\texttt{E$_4$=5}).
